Word Sense Disambiguation\\ \small{Leveraging the NLTK Toolkit and External Tools}

### 2011-04-14 paper/paper.tex
the rest of the paper and cleanup

\section{Overview}
We  chose to implement our system in Python, and chose to use the machine learning algorithms in the NLTK toolkit which is also Python-based.

@@

\subsection{Data details}
The  NLTK has a corpus reader  that imports Senseval-2 files. The Senseval-3  file format is slightly  different which necessitated a few changes. We separated the training  and test files by word into .pos files to match  what the existing  parser looked for. In the interest of time we modified  the training and  test files to contain a single tagged word per chunk  (but still  allowing for multiple senses) again to match the Senseval-2  format.

\subsection{Selecting optimal feature combination}
\newcommand\ward{forward} %or maybe backwards
We used a \ward stepwise approach to indentifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$ combinations.
We then used backwards stepwise regression to identify feature combinations
with a small number of features but with high performance

@@

The NLTK implements a naïve bayes classifier which chooses the best sense given a dictionary of features and assigns confidence probabilities to the chosen sense. Naive bayes does assume an independence of features. The probability of each feature contributes to the overall choice.

@@

From the parser we rendered several features:
\begin{description}

@@

Dependency parsing is fairly slow, requiring about a second per instance to run on a computer running an Atom D525 processor (which is fairly underpowered, but modern).  To help alleviate this problem, we wrote a system that saves parses for each context, and only calls the parser if it observes a new parse, so we could reuse parsing information across runs of our program.  Because this file is several hundred megabytes in size, it is not included in our submitted code.

\subsection{Collocation}

@@

\subsection{Cooccurrence}
Looks through the training data and compiles a list of most common words for each item in the data. When the test data is run, the feature looks for the created vector and uses that instead of trying to make a new one. Next the extractor takes a window of words around the head word and generates the 0, 1 tuple of features for inclusion to the dictionary. This word list is used as the keys of the feature vector when using the test data as a list of words that may cooccur with each item in the data. The value associated with each key in the feature vector starts out as 0 (for occurring 0 times in the window near a test word) If a key in the feature vector does occur with a word in the test data, the value associated with the key changes to 1 even if the key occurs more than once.  This is necessary because the machine learning algorithm we used does not attempt to analyze the keys beyond simple equality, so higher numbers than one could have masked the possibly more important distinction between 0 and not-0. In other words, keeping track of the instances of occurrence vs clustering them into 2 groups reduces the sparsity of the data and allows for a higher chance of accurate classifications based on the limited training data that we have. This is an especially important consideration since the feature vectors that select the most common words by looking through all the instances of unique lexical items in the training data are unsupervised.

\section{Methods}
\newcommand\few{few}
We ran the system with several hundred different combinations of features over all the words in the input set.

@@

\subsubsection{Stepwise regression}
We ran backward stepwise regression with the F-measure from the coarse-grained scores
regressed on predictors representing the feature combination.
The saturated model contained the following predictors and all interactions
\begin{itemize}
\item Classifier (naive Bayes or decision list)
\item Collocation window length (0--8)
\item Cooccurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}
The predictor combination yeilding the lowest value of the Akaike information criterion
included the following parameters.
\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Coocurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$cooccurrence interaction
\end{itemize}
This does not include bootstrap iterations, indicating that changes
in bootstrap iterations does not substantially change the F-measure.

%Discuss the coefficients in the regression table
Call:
lm(formula = f ~ classifier + collocation + cooccurrence + base_word +     dependency_parsing + classifier:collocation + classifier:cooccurrence +     collocation:cooccurrence + collocation:dependency_parsing +     cooccurrence:dependency_parsing + collocation:cooccurrence:dependency_parsing,     data = plot.step.import())

Coefficients:
\begin{tabular}{l r}
                                (Intercept) & 0.5557355 \\
                                classifiert & 0.0444331  \\
                                 collocation & 0.0030636  \\
                              cooccurrence1 & 0.0276413 \\
                                 base_word1 & 0.0124008  \\
                         dependency_parsing & -0.0019423 \\
                     classifiert:collocation & -0.0132109 \\
                  classifiert:cooccurrence1 & -0.0271825 \\
                   collocation:cooccurrence1 & -0.0007417 \\
              collocation:dependency_parsing & -0.0003464 \\
           cooccurrence1:dependency_parsing & -0.1611043 \\
collocation:cooccurrence1:dependency_parsing & 0.0147078  \\
\end{tabular}

@@

\subsubsection{Common Words in Context}
An idea for a feature extractor was to take the most common words in a context of a lexeme and use those words as features. Based off of the cooccurrence code, this would generate a vector from the training data, find the most common words from the test data, and then run the collected data through the classifier.

@@

\subsubsection{Part of Speech Tagging For Target Words}
We built a part-of-speech tagger and trained it on the Brown corpus. We planned on using it to tag words, but the target words were already grouped by part of speech, (activate.\texttbf{v}) so we did not use it.

>>>>

\section{Overview}

We  chose to implement our system in Python, and chose to use the machine
learning algorithms in the NLTK toolkit which is also Python-based.

@@

\subsection{Data details}

The  NLTK has a corpus reader  that imports Senseval-2 files. The Senseval-3
file format is slightly  different which necessitated a few changes. We
separated the training  and test files by word into .pos files to match  what
the existing  parser looked for. In the interest of time we modified  the
training and  test files to contain a single tagged word per chunk  (but still
allowing for multiple senses) again to match the Senseval-2  format.

\subsection{Selecting optimal feature combination}
\newcommand\ward{forward} %or maybe backwards

We used a \ward stepwise approach to indentifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$
combinations.  We then used backwards stepwise regression to identify feature
combinations with a small number of features but with high performance

@@

The NLTK implements a naï bayes classifier which chooses the best sense given a
dictionary of features and assigns confidence probabilities to the chosen
sense. Naive bayes does assume an independence of features. The probability of
each feature contributes to the overall choice.

@@

From the parser we rendered several features:

\begin{description}

@@

Dependency parsing is fairly slow, requiring about a second per instance to run
on a computer running an Atom D525 processor (which is fairly underpowered, but
modern).  To help alleviate this problem, we wrote a system that saves parses
for each context, and only calls the parser if it observes a new parse, so we
could reuse parsing information across runs of our program.  Because this file
is several hundred megabytes in size, it is not included in our submitted code.
\subsection{Collocation}

@@

\subsection{Cooccurrence}

Looks through the training data and compiles a list of most common words for
each item in the data. When the test data is run, the feature looks for the
created vector and uses that instead of trying to make a new one. Next the
extractor takes a window of words around the head word and generates the 0, 1
tuple of features for inclusion to the dictionary. This word list is used as the
keys of the feature vector when using the test data as a list of words that may
cooccur with each item in the data. The value associated with each key in the
feature vector starts out as 0 (for occurring 0 times in the window near a test
word) If a key in the feature vector does occur with a word in the test data,
the value associated with the key changes to 1 even if the key occurs more than
once.  This is necessary because the machine learning algorithm we used does not
attempt to analyze the keys beyond simple equality, so higher numbers than one
could have masked the possibly more important distinction between 0 and not-0.
In other words, keeping track of the instances of occurrence vs clustering them
into 2 groups reduces the sparsity of the data and allows for a higher chance of
accurate classifications based on the limited training data that we have. This
is an especially important consideration since the feature vectors that select
the most common words by looking through all the instances of unique lexical
items in the training data are unsupervised.

\section{Methods}
\newcommand\few{few}

We ran the system with several hundred different combinations of features over
all the words in the input set.

@@

\subsubsection{Stepwise regression}

We ran backward stepwise regression with the F-measure from the coarse-grained
scores regressed on predictors representing the feature combination.  The
saturated model contained the following predictors and all interactions

\begin{itemize}
\item Classifier (naive Bayes or decision list)
\item Collocation window length (0--8)
\item Cooccurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}

The predictor combination yeilding the lowest value of the Akaike information criterion
included the following parameters.

\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Coocurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$cooccurrence interaction
\end{itemize}

This does not include bootstrap iterations, indicating that changes
in bootstrap iterations does not substantially change the F-measure.

%Discuss the coefficients in the regression table
%Call:
%lm(formula = f ~ classifier + collocation + cooccurrence + base_word +     dependency_parsing + classifier:collocation + classifier:cooccurrence +     collocation:cooccurrence + collocation:dependency_parsing +     cooccurrence:dependency_parsing + collocation:cooccurrence:dependency_parsing,     data = plot.step.import())

Coefficients:

\begin{tabular}{l r}
                                (Intercept) & 0.5557355 \\
                                classifiert & 0.0444331  \\
                                 collocation & 0.0030636  \\
                              cooccurrence1 & 0.0276413 \\
                                 base word1 & 0.0124008  \\
                         dependency parsing & -0.0019423 \\
                     classifiert:collocation & -0.0132109 \\
                  classifiert:cooccurrence1 & -0.0271825 \\
                   collocation:cooccurrence1 & -0.0007417 \\
              collocation:dependency parsing & -0.0003464 \\
           cooccurrence1:dependency parsing & -0.1611043 \\
collocation:cooccurrence1:dependency parsing & 0.0147078  \\
\end{tabular}

@@

\subsubsection{Common Words in Context}

An idea for a feature extractor was to take the most common words in a context
of a lexeme and use those words as features. Based off of the cooccurrence code,
this would generate a vector from the training data, find the most common words
from the test data, and then run the collected data through the classifier.

@@

\subsubsection{Part of Speech Tagging For Target Words}

We built a part-of-speech tagger and trained it on the Brown corpus. We planned
on using it to tag words, but the target words were already grouped by part of
speech, (activate.\textbf{v}) so we did not use it.
